gazeHMM: Parameter recovery simulation

### 2020-05-10 simulation/preregistration_parameter_recovery.Rmd
update parameter recovery

The HMM is implemented in R (CITE) using the package depmixS4 (CITE). An expectation-maximization algorithm estimates the parameters given the hidden states, which are estimated through the Viterbi algorithm (CITE). The parameters of each response model (except for the uniform distribution) are estimated through maximum likelihood with a spectral projected gradient method (CITE) and Barzilai-Borwein step lengths (CITE) as implemented in the package BB (CITE).

@@

The simulation study will be divided into four parts. In the first part, we will vary the parameters of the HMM. For models with $k \in \{2, 3, 4\}$ states, $q \in \{10, 15, 20\}$ parameters will be varied respectively. For each parameter, the HMM will generate 100 data sets with $N = 2500$ samples and the parameter varied in a specified interval in equidistant steps. This will result in $100 \times (10+15+20) = 4500$ recoveries. Only one parameter will be varied at once, the other parameters will be set to their fixed values. We will not manipulate the initial state probabilities because these are usually irrelevant in the context of eye movement classification. For the transition probabilities, we will only vary the probabilities for staying in the same state (diagonals of the tranistion matrix) to reduce the complexity of the simulation. The left probability mass will be split evenly between the probabilites for switching to a different state (per row of the transition matrix). Moreover, we will not modify the mean parameters of the von-Mises distribution: As location parameters, they do not alter the shape of the distribution and they are necessary features for the HMM to distinguish between different states.

@@

In the second part, we will vary the sample size of the generated data and the amount of noise added to it. The model parameters will be set to their fixed values. For models with $k \in \{2, 3, 4\}$ states and sample sizes of $N \in \{500, 2500, 10000\}$, we will generate 100 data sets ($100 \times 3 \times 3 = 900$ recoveries). We chose samples sizes roughly corresponding to small, medium, and large eye-tracking data sets of a single participant and trial. To simulate measurement error, we will add Gaussian white noise with $\mu = 0$ and $\sigma \in [1,5]$ varying between data sets to velocity and acceleration. Values that decrease to zero or below will be set to 0.001. To angle, we will add white noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ varying in equidistant steps between data sets.

>>>>

The HMM is implemented in R (CITE) using the package depmixS4 (CITE). An expectation-maximization algorithm (with a maximum of 5000 iterations) estimates the parameters given the hidden states, which are estimated through the Viterbi algorithm (CITE). The parameters of each response model (except for the uniform distribution) are estimated through maximum likelihood with a spectral projected gradient method (CITE) and Barzilai-Borwein step lengths (CITE) as implemented in the package BB (CITE).

@@

The simulation study will be divided into four parts. In the first part, we will vary the parameters of the HMM. For models with $k \in \{2, 3, 4\}$ states, $q \in \{10, 15, 20\}$ parameters will be varied respectively. For each parameter, the HMM will generate 100 data sets with $N = 2500$ samples and the parameter varied in a specified interval in equidistant steps. This will result in $100 \times (10+15+20) = 4500$ recoveries. Only one parameter will be varied at once, the other parameters will be set to their fixed values. We will not manipulate the initial state probabilities because these are usually irrelevant in the context of eye movement classification. For the transition probabilities, we will only vary the probabilities for staying in the same state (diagonals of the tranistion matrix) simultaneously to reduce the complexity of the simulation. The left probability mass will be split evenly between the probabilites for switching to a different state (per row of the transition matrix). Moreover, we will not modify the mean parameters of the von-Mises distribution: As location parameters, they do not alter the shape of the distribution and they are necessary features for the HMM to distinguish between different states.

@@

In the second part, we will vary the sample size of the generated data and the amount of noise added to it. The model parameters will be set to their fixed values. For models with $k \in \{2, 3, 4\}$ states and sample sizes of $N \in \{500, 2500, 10000\}$, we will generate 100 data sets ($100 \times 3 \times 3 = 900$ recoveries). We chose samples sizes roughly corresponding to small, medium, and large eye-tracking data sets of a single participant and trial. To simulate measurement error, we will add Gaussian white noise with $\mu = 0$ and $\sigma \in [1,5]$ varying between data sets to velocity and acceleration. Values that decrease to zero or below will be set to 0.001. To angle, we will add white noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ varying between data sets. $\sigma$ and $\kappa$ will vary simultaneously in equidistant steps in their intervals.
